---
title: "Coursera <br> Developing Data Products <br> Week 4 Project"
author: "Paolo Guderzo"
date: "May 11th, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This project corresponds to the final assignment of the **Developing Data Products** Coursera course by Johns Hopkins University.
The main goal of this project is to create a shiny application to be deployed on a shiny server. The link for the application is https://duplo59.shinyapps.io/myApp/. 
The code is available here: https://github.com/Duplo59/datasciencecoursera/tree/master/Course_09-Developing_Data_Products

In order to do the project, I used the 'USArrests' dataset. The application performs a brief analysis on this dataset using some barplots and scatterplots. The user can choose the analysis he desires using four radio buttons.

## Dataset overview

The dataset contains violent crime data (assaults, rapes and murders - data per 100,000 residents) collected in USA in 1973 (50 observations on 4 variables). Besides, the dataset contains, for every observation, the percentage of urban population.

```{r, datasetOverview}
head(USArrests)
```

## UI Code

```{r, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
library(shiny)
shinyUI(fluidPage(
  titlePanel("A brief analysis about US Arrests in 1973"),
  sidebarLayout(
    sidebarPanel(
        radioButtons("modelType",
                     label = "Which kind of analysis do you want?",
                     choices =  list("Correlation analysis" = 1,
                                     "Analysis by US State" = 2,
                                     "Murders vs Assaults" = 3,
                                     "Rapes vs Assaults" = 4),
                     selected = 1)
    ),
    mainPanel(
      plotOutput("varPlot")
    )
  )
))
```

## Some server code

```{r, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
# Based on the choice done by the user, a scatter plot 
# with a regression line is displayed
shinyServer(function(input, output) {
    reactive({input$modelType
    })
    observe({
        if (input$modelType == 1) {
            output$varPlot <- renderPlot({ggpairs(df, 
                    lower = list(continuous = "smooth"),
                    title = "Some insights abouts correlations") +
                    theme_bw() + 
                    theme(plot.title = element_text(hjust = 0.5)) +
                    theme(plot.title = element_text(color="black", 
                    size=16, face="bold.italic"))
                    })
        } else if (input$modelType == 2) {
            output$varPlot <- renderPlot({ggplot(df_State, 
                aes(x = State, y = Number)) +
                geom_bar(stat="identity", aes(fill = Crime), 

```


